---
title: "WVS Injunction: RST Map"
author: "Mairin Deith, m.deith@oceans.ubc.ca"
date: "`r paste0('Last updated on ', format(Sys.time(), '%Y-%b-%d'))`"
output:
  html_document:
    number_sections: false
    theme: flatly
---

# Background 

Rotary screw traps (RSTs) are used to monitor downstream migration of salmonids throughout the Willamette Valley basin. 
The locations of the RSTs, years of operation, and operator IDs are described in the map below, and in an interactive data table. 

# Map of all RST locations (including those where UBC does not have data)

<span style="font-size:120%">This map includes several locations which were estimated or assumed from previous years. Scroll further down the page to see a map of <strong>only</strong> verified locations.</span>  

Below, you can interact with the map using your cursor. 
Hovering over a project, river, reservoir, or other feature will open help-text that explains what is under your cursor.
You can also click-and-drag over areas of the map to zoom in.
Double-click to zoom back out. 

Focal subbasins are purple, and focal projects and reservoirs are gold. 
Pink circles represent pre-injunction RST sites, blue triangles represent post-injunction sites.
Black circles are USGS gages used during analysis (e.g., to verify hydrological data reported at USACE projects, during trap efficiency analysis, and as covariates in statistical models).

<br>

```{r map-it-estimated, echo = F, message = F, warning = F, out.width = "100%", out.height = "100%"}
setwd("C:/Users/mdeith/OneDrive - UBC/General/Injunction Analysis/")
setwd("injunction_timeline/code/WVSInjunctionAnalysis.io")

# Uses the outputs from the "add_rst_to_basemaps.R" script 

library(ggplot2)
library(plotly)

overview_map_ggplotly <- readRDS("data/2024_07_17_rst_map_compiled_estimatedlocs_gages.Rds")

overview_map_ggplotly %>%
  plotly::layout(
    title = "All RST locations (incl. estimated)",
    legend = list(
        x = 0, 
        y = 1, 
      orientation = 'h', 
      font = list(size = 14)),
      autosize = T, height = 800, width = "100%")
```

# Data table

Here is the data used to populate the map above, but in table form.
Using the inputs at the top of each column, you may filter the data to specific years, locations, etc.

```{r table-it, echo = F, message = F, warning = F, out.width = "100%"}
library(DT)
library(readxl)

rst_raw <- read_xlsx("data/Summary of available RST data_estimatesForBlanks_20240717.xlsx", 
    sheet = "Trap Locations") %>%
    # The "notes" column has no header, fix
    rename(Notes = `...12`)
    
datatable(rst_raw %>%
  mutate(
    `Data with UBC` = factor(`Data with UBC`, levels = c("Yes", "No")), 
    Operator = factor(Operator),
    Subbasin = factor(Subbasin),
    Trap = factor(Trap),
    `Injunction period` = factor(`Injunction period`, levels = c("Pre", "Post")),
    `Above or Below` = factor(`Above or Below`, levels = c("Above", "Below")),
    Year = factor(Year),
    Location = factor(Location)
  ) %>%
  select(-c(`UTM (NAD 83)`, Lat, Lon)),
  filter = list(position = "top", clear = FALSE),
  rownames = FALSE, 
  width = "100%"
  ) %>%
  DT::formatStyle(columns = c(3, 9), fontSize = '75%') %>%
  DT::formatStyle(columns = c(1,2,4,5,6,7,8), fontSize = '85%')
```

# Map of complete locations only

Below, you can interact with the map using your cursor. 
Focal subbasins are purple, and focal projects and reservoirs are gold. 
Pink circles represent pre-injunction RST sites, blue triangles represent post-injunction sites.

Hovering over a project, river, reservoir, or other feature will open help-text that explains what is under your cursor.
You can also click-and-drag over areas of the map to zoom in.
Double-click to zoom back out. 

```{r map-it-complete, echo = F, message = F, warning = F, out.width = "100%", out.height = "100%"}
setwd("C:/Users/mdeith/OneDrive - UBC/General/Injunction Analysis/")
setwd("injunction_timeline/code/WVSInjunctionAnalysis.io")

# Uses the outputs from the "add_rst_to_basemaps.R" script in the "supp_code" folder of the website page

library(ggplot2)
library(plotly)

overview_map_ggplotly <- readRDS("data/2024_06_26_v2_rst_map_compiled_yrranges.Rds")

overview_map_ggplotly %>%
  plotly::layout(
    title = "Filtered RST locations",
    legend = list(
        x = 0, 
        y = 1, 
      orientation = 'h', 
      font = list(size = 14)),
      autosize = T, height = 800, width = "100%")
```
